---
title: "Beyond the Prompt"
subtitle: "Programming as a Pathway to Statistical Thinking"
author: "Mine Çetinkaya-Rundel"
date: today
date-format: long
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: c/t
    progress: true
    transition: fade
    background-transition: fade
    center-title-slide: true
    title-slide-attributes:
      data-background-color: "#496275"
    highlight-style: github
    code-line-numbers: true
    fig-align: center
    fig-width: 8
    fig-height: 4.5
execute:
  echo: true
  warning: false
  message: false
---

## Slide 1

::: notes
Thank you so much for the invitation to speak today. I am so grateful for this invitation and so excited to open the conference. Especially because this means that after this hour I get to sit back and listen to all of you talk about the exciting work you are doing in your classrooms. I am really looking forward to that.

Option 1: You know what else I've been sitting back and watching lately, and sometimes enjoying and sometimes spiraling into existential panic about? AI tools crushing homework assignments I've meticulously, painstakingly, and lovingly designed for my students. Let's take a look...

Option 2: I've just celebrated my 15th year of teaching at Duke. And you know how academic rewards years of experience -- with more service :) Last year I was appointed the inaugural Director of First-Year Experience at Duke. And yes, it's a big admin role, and that might not sound super exciting, but not only is it a role that allows me to think deeply about how we support students in their first year, but it has surprisingly given me many opportunities for real data science tasks.
:::

## Slide 2: The video

{{< video videos/open-secrets-posit-ai.mov >}}

::: notes
Walk through Posit AI video...

I recently had a lovely visit to UC Santa Cruz and one of the talks I gave there was about AI tools and learning, and I showed a similar demo -- different example but same idea. An audience member remarked, "Was this meant to be rage-bait?". So, yeah. I figured some of you might still be jetlagged from long travels. Hopefully this rage-bait woke you up :)
:::

## Slide 3: The premise {background-color="#496275"}

TO DO: Black background, large block white letters, yes.

::: notes
This brings us to the premise of this talk. Now that AI tools can generate, what I would call functional code, scaffold entire reports, and debug pipelines faster than most students can type, should we still ask students to learn to do these things themselves? The answer is yes. And not for the sake of learning syntax (that was never the final goal anyway), but because it teaches them ways of formulating and solving problems with programmatically, and that's a skill that will serve them well, whether they're writing artisinal code or prompting a model to do it for them.
:::

## Slide 4: Point 1 - Look at data

> From "The Art of Statistics"
>
> "There is no substitute for simply looking at data properly."
> 
> David Spiegelhalter
>
> Professor, University of Cambridge

## Slide 5: 

```{r}
#| echo: false
#| message: false
library(tidyverse)
pac_raw <- read_csv("open-secrets-parens/pac-raw.csv")
pac <- read_csv("open-secrets-parens/pac.csv")
```


## "Cleaned" data

```{r}
pac |>
  filter(if_any(c(total, dems, repubs), ~ .x < 0))
```

## Raw data

Negative values!

```{r}
pac_raw |>
  filter(if_any(c(Total, Dems, Repubs), ~ str_detect(.x, "\\(")))
```

## Cleaned data

```{r}
pac |>
  filter(pac_name == "BP (BP)")
```

::: notes
The open parens are missed by AI. Humans who know what they mean wouldn't miss them, if they saw them. Humans who are curious 
:::

# Why programming still matters {background-color="#496275"}

## When code is free, mastery is the differentiator

The valuable skills shift from *producing* code to *judging* it:

::: {.incremental}
- **Read** a generated script and assess whether it does what you think
- **Recognize** when a pipeline quietly produces the wrong answer
- **Structure** an analysis so every step can be followed and verified
:::

. . .

::: {.key-point}
A plausible-looking answer is now only a prompt away. The pedagogical challenge is rewarding genuine understanding over fluent generation.
:::

## The quietly wrong answer

```{r}
#| label: silent-bug
#| echo: true
library(dplyr)

sales <- tibble(
  region = c("N", "N", "S", "S", NA),
  revenue = c(100, 150, 200, NA, 90)
)

sales |>
  group_by(region) |>
  summarize(mean_rev = mean(revenue))
```

. . .

Looks fine — until you notice the `NA` group and the dropped observation.
Would a student *catch* this in AI-generated code?

# Three prompts, three quiet problems {background-color="#496275"}

## Prompt 1: "Join these two tables" {.smaller}

Two tables whose keys are duplicated in ways we might not expect:

```{r}
#| echo: true
library(tidyverse)

df_1 <- tibble(id = c(1:50, 51, 51, 51, 52:100))
df_2 <- tibble(id = c(1:20, 21, 21, 21, 21, 22:200))

df_1 |>
  left_join(df_2, join_by(id))
```

::: {.notes}
Set this up as an everyday task — you just want to combine two tables on `id`.
Point out that the ids look innocuous, but 51 is repeated three times in the
first table and 21 is repeated four times in the second. Note that even the
plain `left_join` emits a many-to-many warning here. Ask the room: how many of
your students would read that warning versus scroll past it?
:::

## The warning *is* the lesson {.smaller}

The AI's fix — silence it:

```{r}
#| echo: true
#| warning: false
df_1 |>
  full_join(df_2, join_by(id), relationship = "many-to-many") |>
  nrow()
```

. . .

::: {.key-point}
`relationship = "many-to-many"` makes the warning disappear, but the warning
was never a syntax problem — it flags a **data** problem. What if these were
two different people who happen to share an `id`?
:::

. . .

*Does AI reliably spot this? How do we teach students to look?*

::: {.notes}
This is the crux of the joins example. The AI's instinct is to make the warning
go away by declaring the relationship many-to-many — which is syntactically
valid and produces output. But the warning was never about syntax; it was
dplyr telling you the keys aren't unique. Bring it home with the "two different
people sharing an id" scenario: no error will ever fire, the join just silently
doubles rows. The teaching goal is the habit of asking *why* a warning appeared
before dismissing it.
:::

## Prompt 2: "Fit a Poisson regression" {.smaller}

The AI loads the data, fits the model, and reports tidy estimates:

```{r}
#| echo: true
library(broom)

df <- read_csv("data/poisson.csv", show_col_types = FALSE)
df$x2 <- factor(df$x2)

fit <- glm(y ~ x1 + x2, data = df, family = poisson(link = "log"))
tidy(fit, conf.int = TRUE)
```

::: {.notes}
The AI does everything the prompt asked: reads the file, factors the categorical
predictor, fits a Poisson GLM, and hands back clean coefficient estimates with
confidence intervals. Emphasize that nothing here looks wrong — the output is
tidy and confident. A student could paste this straight into a report. Set up
the tension: the question is whether the model is actually appropriate, which
the estimates alone won't tell you.
:::

## A diagnostic worth reading {.smaller}

To its credit, the AI even checks dispersion:

```{r}
#| echo: true
sum(residuals(fit, type = "pearson")^2) / df.residual(fit)
```

. . .

A dispersion of ~8.4 (≫ 1) signals **strong overdispersion** — Poisson standard
errors and p-values are unreliable.

. . .

::: {.key-point}
The estimates *look* authoritative. Whether you trust them depends on reading
a diagnostic the prompt never asked for.
:::

::: {.notes}
Dispersion near 8.4 is the punchline: the Poisson assumption that the variance
equals the mean is badly violated, so the tidy standard errors and p-values from
the previous slide are unreliable. Credit the AI for volunteering this check —
but note the student still has to know what the number means. This is the
difference between generating output and interpreting it. Mention negative
binomial or quasi-Poisson as the natural next step.
:::

## Without AI, I'd plot first {.smaller}

```{r}
#| echo: true
#| fig-height: 4
ggplot(df, aes(x = x1, y = y, color = x2)) +
  geom_point()
```

The picture invites questions the default model missed — like whether `x1` and
`x2` **interact**.

::: {.notes}
Contrast your instinct with the AI's. Before fitting anything, you look at the
data — and the plot immediately raises questions the additive model can't
answer, most obviously whether the two groups have different slopes. The point
isn't that AI can't make this plot; it's that a fluent modeler leads with
exploration, while the prompt-driven workflow jumped straight to a model. This
is "thinking with data" in action.
:::

## Prompt 3: "disp vs mpg with a regression line" {.smaller}

```{r}
#| echo: true
#| fig-height: 4
ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x)
```

::: {.notes}
Start benign: a clear, correct scatter with a linear fit. The AI nailed it.
Use this as the baseline before the follow-up prompt introduces ambiguity on the
next slide.
:::

## "Now limit to displacement below 200" {.smaller}

```{r}
#| echo: true
#| fig-height: 3.5
mtcars |>
  filter(disp < 200) |>
  ggplot(aes(x = disp, y = mpg)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x)
```

::: {.key-point}
The AI *filtered the data* — which changes the fitted line. Did we mean to
subset the data, or just zoom the view? It never asked.
:::

::: {.notes}
This is the subtlest of the three. "Limit to below 200" is genuinely ambiguous:
you might want to restrict the analysis, or you might just want to zoom the axis
(`coord_cartesian(xlim = ...)`) while keeping the full fit. The AI silently
chose the first, which re-estimates the regression on a subset and changes the
line. A careful analyst would have asked which one you meant — because it
changes the answer. The lesson: the interesting work is clarifying intent, not
producing code.
:::

## Thinking with data

Learning modern workflows teaches students to:

::: {.incremental}
- ask **sharper questions**
- notice what a dataset **can and cannot** answer
- build habits of mind for reasoning about **uncertainty** and **modeling**
:::

. . .

These are the foundations of statistical thinking.
The code is how we *practice* them.

# Reproducibility as integrity {background-color="#496275"}

## Not a technical nicety

::: {.incremental}
- Reproducibility is a **cornerstone of scientific integrity**, not a chore
- Tools that enforce it — like **version control** — turn out to be
  essential for working productively *with* AI, too
- A tracked, reviewable history is how you trust a workflow you didn't
  type yourself
:::

## A verifiable workflow

```r
# Every step is legible, ordered, and re-runnable
penguins |>
  filter(!is.na(body_mass_g)) |>          # explicit about missingness
  group_by(species, sex) |>
  summarize(
    n = n(),
    mean_mass = mean(body_mass_g),
    .groups = "drop"
  ) |>
  arrange(desc(mean_mass))
```

::: {.key-point}
If you can read it, you can review it. If you can review it, you can trust it.
:::

# Teaching in the AI era {background-color="#496275"}

## Reframing programming instruction

Not syntax acquisition — a **pathway to statistical thinking**.

Design assignments and assessments for a moment when a plausible answer
is a prompt away:

::: {.incremental}
- tasks that reward **verification**, not just generation
- prompts where the *interesting* work is interpreting the output
- assessment of **reasoning about data**, not recall of function names
:::

## The goal {.center background-color="#496275"}

Not to keep AI out of the classroom —

. . .

but to produce graduates who can do more than

::: {.r-fit-text}
write a good prompt and hope for the best.
:::

## Thank you {background-color="#496275"}

**Beyond the Prompt: Programming as a Pathway to Statistical Thinking**

Mine Çetinkaya-Rundel

Questions and discussion welcome.
