## ICOTS talk code examples

```{r}
library(tidyverse)
```

### Joins

```{r}
df_1 <- tibble(id = c(1:50, 51, 51, 51, 52:100))
df_2 <- tibble(id = c(1:20, 21, 21, 21, 21, 22:200))
```

#### Joins - Without AI

```{r}
df_1 |>
  left_join(df_2, join_by(id))
```

#### Joins - With AI 

> Q: Join df_1 and df_2.

> A:

```{r}
# Full join (allow many-to-many)
df_joined <- df_1 |>
  full_join(df_2, join_by(id), relationship = "many-to-many")

df_joined
```

**What happened?** Avoiding the warning by explicitly dealing with it (or suppressing it?) with justifying why. The warning exists not because of a syntax issue but because of a potentially problematic data issue.

> Q: Can you explain what it means to allow many-to-many?

> A: Many-to-many means the join key is not unique in either table, so each matching key produces a Cartesian product of matching rows (rows multiply). dplyr warns by default because this often indicates a mistake; setting relationship = "many-to-many" tells dplyr you expect that behavior and to proceed.

**What happened?** Not a bad answer, and gives an example to check as well. But learner needs to ask for it.

#### Joins - Further thoughts

What if we have data with matching names but of clearly different people (e.g., two survey respondents with the same name but different ages and zip codes) and no ID column. 

- Does AI reliably spot issues like this? 

- How do we teach students to check for and handle these issues?

### Modeling - With AI

> Q: There is a data file called poisson.csv in the data folder of count observations along with two features - one numerical and the other categorical. Fit a Poisson regression model to these data with R and evaluate its fit. Report the estimates for the regression coefficients you obtain and interpret them in the context of the data.

> A: I'll load the CSV, inspect column names, fit a Poisson GLM (log link), compute coefficient estimates, IRRs, dispersion and a deviance test, and return a summary object.

```{r}
# Load data, inspect, fit Poisson GLM, compute diagnostics
library(tidyverse)
library(broom)

df <- readr::read_csv("data/poisson.csv")
colnames(df)

# Assume columns are: count, numeric, categorical in that order; adapt automatically
cn <- colnames(df)
count_col <- cn[1]
num_col <- cn[2]
cat_col <- cn[3]

# Ensure categorical is a factor
df[[cat_col]] <- factor(df[[cat_col]])

# Build formula and fit
form <- as.formula(paste(count_col, "~", num_col, "+", cat_col))
fit <- glm(form, data = df, family = poisson(link = "log"))

# Summaries and diagnostics
coef_table <- broom::tidy(fit, conf.int = TRUE)
irrs <- broom::tidy(fit, conf.int = TRUE) |>
  mutate(
    estimate_IRR = exp(estimate),
    conf.low_IRR = exp(conf.low),
    conf.high_IRR = exp(conf.high)
  )

dispersion <- sum(residuals(fit, type = "pearson")^2) / df.residual(fit)
resid_dev <- deviance(fit)
resid_df <- df.residual(fit)
p_dev <- pchisq(resid_dev, resid_df, lower.tail = FALSE)

result <- list(
  df_head = dplyr::slice_head(df, n = 6),
  colnames = cn,
  formula = form,
  coef_table = coef_table,
  irr_table = irrs,
  dispersion = dispersion,
  residual_deviance = resid_dev,
  residual_df = resid_df,
  deviance_p_value = p_dev,
  aic = AIC(fit)
)

result
```

```{r}
library(tidyverse)
library(broom)

df <- readr::read_csv("data/poisson.csv")
# Inspect
colnames(df)

df$x2 <- factor(df$x2)

# Fit correct Poisson: response is y
fit2 <- glm(y ~ x1 + x2, data = df, family = poisson(link = "log"))

tidy_coef <- broom::tidy(fit2, conf.int = TRUE)
tidy_irr <- tidy_coef |>
  mutate(
    IRR = exp(estimate),
    IRR_conf.low = exp(conf.low),
    IRR_conf.high = exp(conf.high)
  )

dispersion2 <- sum(residuals(fit2, type = "pearson")^2) / df.residual(fit2)
res_dev2 <- deviance(fit2)
res_df2 <- df.residual(fit2)
p_dev2 <- pchisq(res_dev2, res_df2, lower.tail = FALSE)
AIC2 <- AIC(fit2)

list(
  coef = tidy_coef,
  irr = tidy_irr,
  dispersion = dispersion2,
  residual_deviance = res_dev2,
  residual_df = res_df2,
  deviance_p = p_dev2,
  AIC = AIC2
)
```

> Model fitted: glm(y ~ x1 + x2, family = poisson(link="log")).

> Results (estimates, SE, 95% CI, IRR = exp(estimate)):

> - (Intercept): 1.56 (SE 0.064), 95% CI [1.43, 1.68]; IRR = 4.75 [4.18, 5.37]. Interpretation: expected count ≈ 4.75 when x1 = 0 and x2 = reference level (x2 = "a").
> - x1: −0.194 (SE 0.082), 95% CI [−0.355, −0.033]; IRR = 0.824 [0.701, 0.968], p ≈ 0.018. Interpretation: each one‑unit increase in x1 multiplies the expected count by ≈0.824 (≈17.6% decrease); effect is statistically significant.
> - x2 (b vs a): −0.0441 (SE 0.094), 95% CI [−0.229, 0.140]; IRR = 0.957 [0.795, 1.15], p ≈ 0.64. Interpretation: category b has an expected count ≈0.96× that of a (not statistically different from 1).

> Model fit diagnostics:
> - Residual deviance = 658.76 on 97 df (p << 0), dispersion (Pearson) ≈ 8.39 >> 1.
Conclusion: strong overdispersion — the Poisson model substantially underestimates variance; inference may be unreliable.

> Recommendation: rerun with a model that allows extra‑Poisson variation (negative binomial via MASS::glm.nb or glm(..., family = quasipoisson)) and compare AIC / goodness‑of‑fit.

#### Modeling - Without AI

I'd first plot the data!

```{r}
ggplot(df, aes(x = x1, y = y, color = x2)) +
  geom_point()
```

. . .

Then fit a model with an interaction effect.

#### Modeling - With AI continued

> Q: I'll make two plots: a jittered scatter of y vs x1 colored by x2, and a boxplot of y by x2. The code returns both ggplot objects.

> A:

```{r}
library(ggplot2)

# ensure factor
df$x2 <- factor(df$x2)

# jittered scatter: counts vs numeric predictor, colored by category
p1 <- ggplot(df, aes(x = x1, y = y, color = x2)) +
  geom_jitter(height = 0.2, width = 0, alpha = 0.6) +
  labs(title = "Counts vs x1", x = "x1", y = "count", color = "x2") +
  theme_minimal()

# boxplot by category with jittered points
p2 <- ggplot(df, aes(x = x2, y = y, fill = x2)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.6, aes(color = x2)) +
  labs(title = "Counts by x2", x = "x2", y = "count") +
  theme_minimal() +
  theme(legend.position = "none")

list(scatter = p1, boxplot = p2)
```

> Q: Based on the scatterplot above, is the model you fit a good model?

> A: No — the fit is poor.

> - Evidence from the model you already ran:
>   - Pearson dispersion ≈ 8.39 (≫ 1) — strong overdispersion.
>   - Residual deviance = 658.76 on 97 df, p ≪ 1 (≈ 2.3e-84) — model does not fit.

> - Implication: Poisson variance assumption (Var[Y]=E[Y]) is violated; standard errors and p-values are unreliable.

> Next steps (quick): try a quasipoisson or negative‑binomial and inspect residual plots.

> Brief: fit alternatives with

```{r}
# Fit quasipoisson and negative binomial alternatives
fit_qp <- glm(y ~ x1 + x2, data = df, family = quasipoisson(link = "log"))
library(MASS)
fit_nb <- MASS::glm.nb(y ~ x1 + x2, data = df)
```

> Then compare dispersion, residuals, and AIC (nb) or quasi‑likelihood diagnostics.

### ggplot - With AI

> Q: Create a plot of disp vs. mpg for mtcars and plot a linear regression line on top of the points.

> A: Scatter of displacement vs. mpg with a linear regression line (95% CI):

```{r}
library(ggplot2)

p <- ggplot(mtcars, aes(x = disp, y = mpg)) +
  geom_point(alpha = 0.7) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    color = "blue",
    fill = "lightblue",
    se = TRUE
  ) +
  labs(
    title = "Displacement vs. MPG with linear fit",
    x = "Displacement (cu.in.)",
    y = "Miles per gallon (mpg)"
  ) +
  theme_minimal()

p
```

> Q: Can you limit to cars with displacement below 200?

> A: Filter mtcars to disp < 200 and plot disp vs mpg with an lm fit and 95% CI:

```{r}
# R
library(ggplot2)
library(dplyr)

mtcars_small <- mtcars |> filter(disp < 200)

p_small <- ggplot(mtcars_small, aes(x = disp, y = mpg)) +
  geom_point(alpha = 0.7) +
  geom_smooth(
    method = "lm",
    formula = y ~ x,
    color = "blue",
    fill = "lightblue",
    se = TRUE
  ) +
  labs(
    title = "Displacement < 200: MPG vs displacement",
    x = "Displacement (cu.in.)",
    y = "MPG"
  ) +
  theme_minimal()

p_small
```

**What happened?** Code works but no clarification was requested about whether the data or the plot should be limited, which will change the model fit.
