---
title: "Open Secrets: Foreign-Connected PAC Contributions"
subtitle: "Contributions to US political parties from foreign-connected PACs, 2000-2024"
format: html
execute:
  warning: false
  message: false
---

## Background

Only American citizens (and immigrants with green cards) can contribute to
federal politics, but the American divisions of foreign companies can form
political action committees (PACs) and collect contributions from their American
employees. This analysis works with data on contributions to US political
parties from foreign-connected PACs across 13 election cycles between 2000 and
2024. ([Source](https://www.opensecrets.org/political-action-committees-pacs/foreign-connected-pacs/2024))

```{r}
library(tidyverse)
library(janitor)
```

## a. Load and stack the data

Each CSV covers one election cycle and is named with the cycle's year range
(e.g. `Foreign Connected PACs, 1999-2000.csv`). Passing the vector of file paths
to a single `read_csv()` call stacks them, and the `id` argument records which
file each row came from so we can recover the year.

```{r}
files <- list.files("data", full.names = TRUE)

pac_raw <- read_csv(files, id = "file")

pac_raw
```

## b. Clean and transform

We use `janitor::clean_names()` to standardize the column names, then:

- Derive `year` from the file path, using the **ending** year of each election
  cycle (e.g. the 1999-2000 file becomes `2000`).
- Split `Country of Origin/Parent Company` into separate `country_of_origin` and
  `parent_company` columns. A few entries contain extra `/` characters, so we
  merge any surplus pieces back into the parent company.
- Convert the dollar-formatted `total`, `dems`, and `repubs` strings to numbers.
  Some amounts are written in parentheses (e.g. `($3,000)`), an accounting
  convention for **negative** values, so we can't rely on `parse_number()`
  alone — it strips the parentheses and would return a positive number. The
  `parse_dollar()` helper below flags parenthesized strings and negates them.

```{r}
parse_dollar <- function(x) {
  negative <- str_detect(x, "^\\s*\\(")
  amount <- parse_number(x)
  if_else(negative, -amount, amount)
}
```

```{r}
pac <- pac_raw |>
  clean_names() |>
  mutate(
    year = as.integer(str_extract(str_extract(file, "\\d{4}\\.csv$"), "\\d{4}"))
  ) |>
  separate_wider_delim(
    country_of_origin_parent_company,
    delim = "/",
    names = c("country_of_origin", "parent_company"),
    too_many = "merge"
  ) |>
  mutate(across(c(total, dems, repubs), parse_dollar)) |>
  rename(pac_name = pac_name_affiliate) |>
  select(year, pac_name, country_of_origin, parent_company, dems, repubs, total)

pac |>
  slice_head(n = 10)
```

## c. Pivot longer

To make party a variable rather than encoding it in the column names, we pivot
`dems` and `repubs` into a single `party` column with an accompanying `amount`
column.

```{r}
pac_long <- pac |>
  pivot_longer(
    cols = c(dems, repubs),
    names_to = "party",
    values_to = "amount"
  ) |>
  mutate(party = if_else(party == "dems", "Democrat", "Republican"))

pac_long |>
  slice_head(n = 10)
```

## d. UK contributions by cycle and party

For each election cycle, we sum contributions to each party from PACs whose
`country_of_origin` is the UK. The result has two rows per year, one per party.

```{r}
pac_uk <- pac_long |>
  filter(country_of_origin == "UK") |>
  summarise(total_amount = sum(amount, na.rm = TRUE), .by = c(year, party)) |>
  arrange(year, party)

pac_uk |>
  slice_head(n = 10)
```

## e. Visualization

```{r}
#| fig-width: 8
#| fig-height: 5
#| fig-alt: "Line chart of UK-connected PAC contributions to Democrats and Republicans from 2000 to 2024."

ggplot(pac_uk, aes(x = year, y = total_amount / 1e6, color = party)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(Democrat = "#3A4EB8", Republican = "#D62728")) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  scale_y_continuous(labels = scales::label_dollar(suffix = "M")) +
  labs(
    x = "Election cycle",
    y = "Total contributions",
    color = "Party",
    title = "UK-connected PAC contributions to US parties, 2000-2024"
  ) +
  theme_minimal()
```

**Design choices.** Contributions are a quantity tracked over time, so a line
chart (with points marking each observed cycle) is more appropriate than bars:
it emphasizes the trajectory and makes the two parties' paths easy to compare.
Party is mapped to color using each party's conventional blue/red, which is
intuitive and needs no legend lookup. The x-axis is broken at every election
cycle (every four years for readability), and the y-axis is scaled to millions
of dollars for legibility.

**What it reveals.** Republican-connected UK PAC contributions were consistently
higher than Democratic ones for most of the period, with a notable exception
around the 2008-2010 cycles when Democratic contributions rose sharply and
briefly overtook the Republicans. Both parties saw contributions grow through
the mid-2000s. Republican contributions peaked around 2012-2018 at roughly
\$3.3-3.7M before falling back, while Democratic contributions have generally
trended downward since their 2010 peak. The most recent cycles (2020-2024) show
both parties declining, with the Republican advantage narrowing to a fairly
small gap by 2024.
